Co-citations in context

### 2019-05-16 cocit_r.tex
revision of Intro_gc

\section{Introduction}
Citation and network analysis of  scientific literature reveals information on semantic relationships between publications, collaboration between scientists, and the practice of citation itself~\cite{de_solla_price_networks_1965,newman_structure_2001}. Co-citation, the frequency with which two documents are cited together in other documents~\cite{small_co-citation_1973,marshakova-shaikevich_co-citation_1973}, provides additional insights, including the identification of semantically related documents and fields. Uzzi et al. applied co-citation analysis to 17.9 million publications from the Web of Science (WoS) to identify novel combinations of prior research and to characterize highly cited papers. They analyzed co-citations of references (and journals) within publications, and computed a `z-score' for each journal pair using Monte Carlo simulations based on a random graph model. The set of journal-pair z-scores for each publication was  used to assess  conventionality and novelty, resulting in four categories: high or low conventionality (HC or LC) and high or low novelty (HN or LN), and all four combinations are possible. The authors concluded that highly cited publications across the research literature (and within most subfields) are more likely to be characterized as HC and HN than less highly cited publications.  Key to Uzzi is their random graph model, where  a reference cited in a publication is randomly replaced by another article from the same year as the cited reference, thus producing a random assignment of references to each paper. The model treats potential substitutions as equi-probable, irrespective of disciplinary origin. Accordingly, we hypothesize that analyzing semantically related sets of documents (and limiting substitutions to the citations in the local network) will reduce model misspecification and better simulate citation practice, leading to different and more relevant conclusions.  
We developed a highly efficient algorithm to simulate under a refined random graph model. Our study on three semantically-themed citation networks reveals significantly different patterns of conventionality and novelty between  citation networks and disciplines.

>>>>

\section{Introduction}
Citation and network analysis of  scientific literature reveals information on semantic relationships between publications, collaboration between scientists, and the practice of citation itself~\citep{de_solla_price_networks_1965,newman_structure_2001,Shi:2010:CHI:1816123.1816131,patience_pmid28560354}. Co-citation, the frequency with which two documents are cited together in other documents~\citep{small_co-citation_1973,marshakova-shaikevich_co-citation_1973}, provides additional insights, including the identification of semantically related documents and fields. Co-citations, bibliographic coupling and direct citation have been compared for XX.

In an intriguing study, co-citation analysis of 17.9 million articles and their cited references from the Web of Science (WoS) was used to document novel combinations of prior research and to characterize highly cited papers~\citep{uzzi_atypical_2013}. In this study, $nC2$ combinations of the references in each article were generated and then the reference identifiers were replaced with journal identifiers to generate a journal-pair. Observed frequencies of journal pairs across the dataset were then normalized to expected values generated by a random graph model using a \emph{z-score} (Materials and Methods). Thus, every article was associated with multiple z-scores. For each article, positional statistics of these z-scores were calculated to describe conventionality (high conventionality (HC) if the median z-score for an article was greater than the median z-score of all articles and low conventionality for the inverse (LC).) Similarly with novelty (high novelty (HN) if the tenth percentile of z-scores for an article was less than zero and low novelty (LN) for the inverse). Accordingly, each article was labeled with respect to conventionality and novelty, e.g, HCLN, with all four combinations being possible. The finding of interest was that HCHN articles were twice as likely to be found in highly cited papers (defined as the top 10\% of all papers published in a given year considering the count of citations accumulated in the first 8 years since publication).

Key to the study is the random graph approach, where every reference is randomly replaced by another article published in the same year as the cited reference to generate a null model for comparison. However, the approach inadequately accounts for observed citation behavior~\citep{wallace_lariviere_gingras_2012,moed_measuring_2010,klavans_research_2017} since (i) it treats potential substitutions as equi-probable, irrespective of disciplinary origin (ii) does not consider the frequency with which references are cited in the dataset. In illustration, a reference from musicology can replace one in quantum physics during the simulation process. Similarly, a reference cited over 100 times in a given year is selected with the same probability as a reference cited only once. Accordingly, the expected value calculations used in normalizing observed journal-pair-frequencies can be reasonably questioned on grounds of model mis-specification.

A follow-up study using a smaller dataset of 12 million articles, also suggests that Uzzi and colleagues do not fully account for disciplinary and journal effects~\citep{boyack_vs_uzzi_2014}. While the two studies are conceptually similar, the authors of the latter study used Scopus data and the K50 measurement, a derivative of the $\chi^2$ method, for normalizing observed to expected journal-pair frequencies. Boyack and Klavans note that 'only 64.4\%  of  243  WoS  subject  categories` in the Uzzi study met the criterion of having the highest probability of hit papers in the HCHN category. Further they observed that journals vary widely in terms of size and influence and that 20 journals accounted for nearly 15\% of co-citations in their measurements. Lastly, that three multidisciplinary journals accounted for 9.4\% of all atypical combinations. suggesting strong effects from both disciplines and journals.

Despite different methods used to generate expected values, both Uzzi and Boyack measured observed and expected journal-pair frequencies across disciplines and then examined subsets of publications for disciplinary effects. Since disciplinary origin was not considered when generating expected values in their studies, we hypothesized that analyzing semantically related sets of documents instead of all publications in a literature database would create disciplinary networks. From these, expected values could be calculated, with the effect of reducing model misspecification and better simulating citation practice, in turn leading to different and more relevant conclusions.  We also developed an efficient algorithm that enabled us to conduct a greater number of simulations in a reasonable time. Thus, while Uzzi et al. averaged the results of 10 simulations for each year in their dataset, we performed 1,000 simulations to ensure greater coverage of observed journal pairs. Our study on three semantically-themed citation networks reveals significantly different patterns of conventionality and novelty between  citation networks and disciplines that challenges the conclusion that HCHN articles have a greater probability of being represented in highly cited publications.
